---
title: "Neural network training"
author: "Jose Alberto Arcos"
date: "17 de octubre de 2015"
output: html_document
---

<br><br>

Basic neural network use in R

<br><br>

The neuralnet package seems like a good place to start. There is a simple example [here](http://www.r-bloggers.com/using-neural-networks-for-credit-scoring-a-simple-example/). A more in-depth view of the package can be found in [this paper](http://journal.r-project.org/archive/2010-1/RJournal_2010-1_Guenther+Fritsch.pdf).

The neuralnet package is able to create and train networks with more than one hidden layer, so it is perfect for the kind of problems I want to solve.

<br><br>

I will use the HairEyeColor dataset, which contains info about the frequency of hair and eye color combinations by sex.

```{r}
head(as.data.frame(HairEyeColor))
```

<br><br>

First of all, let's re-structure and plot the data:

```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(magrittr)

data <- as.data.frame(HairEyeColor)
expanded.data <- data.frame("Hair" = character(), 
                            "Eye" = character(), 
                            "Sex" = character())

for(i in 1:nrow(data)){
category <- data_frame("Hair" = rep(data[i, 1], data[i, 4]),
"Eye" = rep(data[i, 2], data[i, 4]),
"Sex" = rep(data[i, 3], data[i, 4]))
expanded.data %<>%
rbind(category)
}

expanded.data %<>%
    mutate(hair.black = as.numeric(Hair == "Black"),
           hair.brown = as.numeric(Hair == "Brown"),
           hair.red = as.numeric(Hair == "Red"),
           hair.blond = as.numeric(Hair == "Blond"),
           eyes.brown = as.numeric(Eye == "Brown"),
           eyes.blue = as.numeric(Eye == "Blue"),
           eyes.hazel = as.numeric(Eye == "Hazel"),
           eyes.green = as.numeric(Eye == "Green"),
           sex.male = as.numeric(Sex == "Male"),
           sex.female = as.numeric(Sex == "Female"))

# Plot data

ggplot(data = as.data.frame(HairEyeColor)) +
geom_jitter(mapping = aes(x = Hair, y = Eye, color = Sex, size = Freq),
position = position_jitter(width = 0.1, height = 0.1),
alpha = 0.8)
```

<br><br>

Now, let's separate training and test sets. I want to predict the eye color of a person given the hair color and the sex.

```{r}
index.training <- sample(x = 1:nrow(expanded.data), size = 500, replace = FALSE)
index.test <- !(1:nrow(expanded.data) %in% index.training)
training <- expanded.data[index.training, ]
test <- expanded.data[index.test, ]
```

<br><br>
And train a neural network to solve the problem.

```{r message=FALSE, warning=FALSE}
library(neuralnet)

nn.eyes.brown <- 
    neuralnet(formula = eyes.brown ~ hair.black + hair.brown + hair.red + 
                  hair.blond + sex.male + sex.female,
              data = training,
              hidden = 1,
              lifesign = "full")
nn.eyes.blue <- 
    neuralnet(formula = eyes.blue ~ hair.black + hair.brown + hair.red + 
                  hair.blond + sex.male + sex.female,
              data = training,
              hidden = 1,
              lifesign = "full")
nn.eyes.hazel <- 
    neuralnet(formula = eyes.hazel ~ hair.black + hair.brown + hair.red + 
                  hair.blond + sex.male + sex.female,
              data = training,
              hidden = 1,
              lifesign = "full")
nn.eyes.green <- 
    neuralnet(formula = eyes.green ~ hair.black + hair.brown + hair.red + 
                  hair.blond + sex.male + sex.female,
              data = training,
              hidden = 1,
              lifesign = "full")

```

<br><br>
Test prediction results:

```{r}
prediction.brown <- 
    neuralnet::compute(nn.eyes.brown, 
                       test %>% dplyr::select(hair.black, hair.brown, hair.red,
                                           hair.blond, sex.male, sex.female))
prediction.blue <- 
    neuralnet::compute(nn.eyes.blue, 
                       test %>% dplyr::select(hair.black, hair.brown, hair.red,
                                       hair.blond, sex.male, sex.female))
prediction.hazel <- 
    neuralnet::compute(nn.eyes.hazel, 
                       test %>% dplyr::select(hair.black, hair.brown, hair.red,
                                       hair.blond, sex.male, sex.female))
prediction.green <- 
    neuralnet::compute(nn.eyes.green, 
                       test %>% dplyr::select(hair.black, hair.brown, hair.red,
                                       hair.blond, sex.male, sex.female))
test$prediction.brown <- prediction.brown$net.result %>% round(2)
test$prediction.blue <- prediction.blue$net.result %>% round(2)
test$prediction.hazel <- prediction.hazel$net.result %>% round(2)
test$prediction.green <- prediction.green$net.result %>% round(2)

test %<>%
    mutate(max.prediction.value = pmax(prediction.green, prediction.hazel, 
                                      prediction.blue, prediction.brown)) %>%
    mutate(prediction.eye = ifelse(prediction.blue == max.prediction.value,
                                   yes = "Blue",
                                   no = NA)) %>%
    mutate(prediction.eye = ifelse(prediction.green == max.prediction.value,
                                   yes = "Green",
                                   no = prediction.eye)) %>%
    mutate(prediction.eye = ifelse(prediction.brown == max.prediction.value,
                                   yes = "Brown",
                                   no = prediction.eye)) %>%
    mutate(prediction.eye = ifelse(prediction.hazel == max.prediction.value,
                                   yes = "Hazel",
                                   no = prediction.eye))
# Show results

test %>% dplyr::select(Sex, Hair, Eye, prediction.eye, max.prediction.value)

accuracy <- test$Eye == test$prediction.eye
accuracy <- sum(accuracy) / length(accuracy)
cat("accuracy = ", accuracy, fill = T)
```